---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
 
---
setwd("C:\\Users\\Audra\\Documents\\GitHub\\RepData_PeerAssessment1")

## Loading and preprocessing the data
###Show any code that is needed to


###### 1. Load the data (i.e. "read.csv())" 

```{r  results=FALSE}

library(dplyr)
library(lubridate)
activity <- read.csv(unz("activity.zip", "activity.csv"))

```
 
###### 2. Process/transform the data (if necessary) into a format suitable for your analysis


## What is mean total number of steps taken per day?


```{r mean total by day}

mean_by_date <- summarise(group_by(activity, date), mean(steps))
names(mean_by_date)[2]<- "mean_total_steps"

```

######For this part of the assignment, you can ignore the missing values in the dataset.

######    Create a new variable activity_no_na and remove values with NA from the steps variable.
```{r removeNA}
activity_no_na <- subset(activity, !is.na(steps ))
```

###### 1 Make a histogram of the total number of steps taken each day
```{r histogram}
sum_by_date <- summarise(group_by(activity, date), sum(steps))
names(sum_by_date)[2]<- "total_steps"
hist(sum_by_date$total_steps,main="Total Number of steps taken each day",xlab="Total Steps")
```

###### 2 Calculate and report the mean and median total number of steps taken per day
```{r mean and median}
sum_by_date <- summarise(group_by(activity_no_na, date), sum(steps))
names(sum_by_date)[2]<- "total_steps"
mean<- mean(sum_by_date$total_steps)
median<-median(sum_by_date$total_steps)
```

The mean is `r mean` and the median is `r median`.

## What is the average daily activity pattern?

######1. Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
```{r timeseries}
average_daily <- summarise(group_by(activity_no_na, interval), mean(steps))
names(average_daily)[2]<- "Average"
plot(average_daily$interval,average_daily$Average, xlab="Interval",ylab="Average Steps",main="5-Minute interval vs Average steps taken", type = "l")
```

######2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r max 5-min interval}
maximum<-max(average_daily$Average)
mi<-subset(average_daily ,Average == maximum)

```

The 5-min interval which contains the maximum number of stes, on average accross all days in the dataset is : `r mi$interval`

## Imputing missing values

###### Note that there are a number of days/intervals where there are missing values (coded as NA). The presence of missing days may introduce bias into some calculations or summaries of the data.


###### 1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)
```{r total number of missing values}
missing <- subset(activity, is.na(steps))
len<-length(missing$steps)
```

The total number of missing values in the dataset is: `r len`.


######2. Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.


######3. Create a new dataset that is equal to the original dataset but with the missing data filled in.
```{r fill missing values}

df <-merge(activity, average_daily, by.x="interval", by.y="interval")
df<-mutate(df, stepsfill= steps)

for (i in 1:length(df$steps))
  if(is.na(df$steps[i])){df$stepsfill[i]=df$Average[i]}

```

######4. Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?
```{r HistogramTotalSteps}
sum_by_date <- summarise(group_by(df, date), sum(stepsfill))
names(sum_by_date)[2]<- "total_steps"
hist(sum_by_date$total_steps,main="Total Number of steps taken each day(No missing variables)",xlab="Total Steps")


mean<- mean(sum_by_date$total_steps)
median<-median(sum_by_date$total_steps)
```
The mean is `r mean` and the median is `r median`. 


## Are there differences in activity patterns between weekdays and weekends?
######For this part the weekdays() function may be of some help here. Use the dataset with the filled-in missing values for this part.


######1. Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.
```{r date factors}

df <- mutate(df,weekday=  wday(as.Date(df$date)) )
df <- mutate(df, weekend = wday(as.Date(df$date)))
df<- mutate(df, check = wday(as.Date(df$date)))

for (i in 1:length(df$weekday))   
      {
        if (df$weekday[i] == 1 | df$weekday[i]== 7)
          { df$weekday[i]=0
            df$weekend[i]=1}
        if( df$weekday[i] ==2 | df$weekday[i]==3 | df$weekday[i]==4| df$weekday[i]==5| df$weekday[i]==6 )
          {df$weekday[i]=1 
           df$weekend[i]=0}
           }   
 df$weekend <- as.factor(df$weekend)
 df$weekday <- as.factor(df$weekday)
```

######2. Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). The plot should look something like the following, which was creating using simulated data:
```{r PanelPlot}
library(lattice)
avg <- summarise(group_by(df,interval,weekday),mean(steps, na.rm = TRUE))
names(avg)[3]<- "Average"

plot(avg$Average,avg$Interval, ylab="Number of steps",xlab="Interval",main="5-Minute interval vs Average steps taken", type ="l")

xyplot(Average ~ interval|weekday,type="l",avg, groups= weekday)

```

###There are differences in the average number of steps taken across weekdays and weekends
